Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-17 paper/main.tex
add comments from Hysan

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query Girard-Hutchinson estimate 
    \begin{equation}
        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

@@

    The second term is first processed with standard matrix-norm inequalities:
    \begin{equation}
        \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F 
        \leq \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F
        \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
        \label{equ:nystrom-proof-processed-tail}
    \end{equation}
    Its higher order moments are therefore bounded by
    \begin{align}
        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
        &\leq \sqrt{k} \cdot \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right] && \text{(using \refequ{equ:nystrom-proof-processed-tail})} \notag \\
        &= \sqrt{k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
        &\leq \sqrt{k} c \cdot \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\sfrac{1}{2}}$)} \notag \\
        &= \sqrt{k} c \cdot \lambda_{k+1} && \text{(from the definition of $\lVert \cdot \rVert _2$)} \notag \\
        &\leq \frac{c}{\sqrt{k}} \Trace(\mtx{B}) && \text{($\lambda_{k+1} \leq \Trace(\mtx{B})/k$)}
        \label{equ:nystrom-proof-tail-bound}
    \end{align}
    Thus, using the triangle inequality for $\mathbb{E}^{\sfrac{k}{2}}$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
    \begin{equation}
        \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
        \leq \frac{1 + c}{\sqrt{k}} \Trace(\mtx{B})
    \end{equation}

    We use Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} to get
    \begin{equation}
        \mathbb{E}^{\sfrac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right] ~\mathrm{d}t% \notag \\
        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
        \leq \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{k}{2}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
    \end{equation}
    from which follows the first statement by replacing $k = \frac{n_{\mtx{\Omega}}}{2}$.

>>>>

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query Girard-Hutchinson estimate 
    \begin{equation}
        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

@@

    The second term is first processed with standard matrix-norm inequalities:
    \begin{equation}
        \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F 
        \leq \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F
        \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
        \label{equ:nystrom-proof-processed-tail}
    \end{equation}
    Its higher order moments are therefore bounded by
    \begin{align}
        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
        &\leq \sqrt{k} \cdot \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right] && \text{(using \refequ{equ:nystrom-proof-processed-tail})} \notag \\
        &= \sqrt{k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(by definition of $\mathbb{E}^{k}$)} \notag \\
        &\leq \sqrt{k} c \cdot \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\sfrac{1}{2}}$)} \notag \\
        &= \sqrt{k} c \cdot \lambda_{k+1} && \text{(from the definition of $\lVert \cdot \rVert _2$)} \notag \\
        &\leq \frac{c}{\sqrt{k}} \Trace(\mtx{B}) && \text{($\lambda_{k+1} \leq \Trace(\mtx{B})/k$)}
        \label{equ:nystrom-proof-tail-bound}
    \end{align}
    Thus, using the triangle inequality for $\mathbb{E}^{\sfrac{k}{2}}$ and inserting \refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud} we obtain
    \begin{equation}
        \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
        \leq \frac{1 + c}{\sqrt{k}} \Trace(\mtx{B}).
    \end{equation}

    As in \cite{kressner-2023-randomized-lowrank}, we can show that the error $\left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]$ is measurable and by the Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities}, we get
    \begin{equation}
        \mathbb{E}^{\sfrac{k}{2}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right] ~\mathrm{d}t% \notag \\
        %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
        %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
        \leq \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
    From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{k}{2}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{1 + c}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
    \end{equation}
    from which follows the first statement by replacing $k = \frac{n_{\mtx{\Omega}}}{2}$.
